A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-23 manuscript.tex
:construction: clustering intro str

\subsection{Feature Clustering}
For feature-based time-series clustering, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible \citep{liao2005}. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment \citep[or similarity measurement for qualitative features; see][]{Aghabozorgi2015}. While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, we will briefly introduce some of the more readily available approaches suitable for most time series feature data. The well-established and readily available clustering approaches can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof \citep[see \tblref{tab:clusterApproaches} for an overview; also see][for a broader review]{jain1999}.

Each of these approaches can be a valuable clustering approach for time series feature data and users will usually have to make an informed decision based on the structure of their data as well as an appropriate weighing of accuracy and efficiency. As an example, under ideal conditions, most approaches are likely to provide a similar cluster solution \citep[e.g., for well-separated groups with little noise and few outliers; e.g., see][]{peng2022}. However, when the shapes of clusters, for example, become more complex in a multi-dimensional space, density-based or hierarchy-based approaches that allow for more bottom-up clustering are likely to be more accurate \citep[e.g.,][]{langfelder2008}. Yet, with higher numbers of participants and features, both density- and hierarchy-based approaches may perform less well and the more efficient centroid-based methods might be more effective \citep[e.g.,][]{jain2010}. Similarly, several of the proposed time series features are statics that are generated from Gaussian distributions and a distribution-based algorithm might be ideally suited to separate such distributions \citep[e.g.,][]{corduas2008}. Yet, in cases where misspecifications are costly, a method with fewer assumptions might be more advisable \citep[e.g.,][]{ankerst1999}.

There is thus, unfortunately, no one-size-fits-all solution to clustering. On the bright side, however, the thriving methodological developments in the clustering literature offer a wide variety of options that offer prospects even for extremely large and complex data sets. We provide a short intuitive explanation for common approaches, together with some of their characteristics and example algorithms in \tblref{tab:clusterApproaches}. For our own illustration, we have chosen that centroid-based k-means clustering. While k-means comes at the expense of high accuracy with more complex cluster shapes, we specifically chose k-means because it is an extremely efficient method that works well with large participant- and feature numbers without making too many restrictive assumptions about the shape of the data \citep{jain2010}. K-means is also well-established within the research community and has been readily implemented in many statistical software packages \citep{hand2005}. Additionally, many of the feature selection methods have specifically been designed for the well-established k-means algorithm \citep[e.g.,][]{boutsidis2010}. As such, the k-means offers a good starting point for many psychological researchers and the method should be generalizable across a relatively wide variety of projects.

>>>>

\subsection{Feature Clustering}
For the actual clustering of the time-series features, the main aim is to organize participants into groups so that the features of participants within a group are as similar as possible, while the features of people in different groups are as different as possible \citep{liao2005}. The crux of clustering is, thus, to have clearly defined and effective measurements of (dis)similarity. Most of the clustering algorithms used today use some form of distance measurement to optimize group assignment \citep[or similarity measurement for qualitative features; see][]{Aghabozorgi2015}. While others have produced excellent overviews of the many clustering approaches available \citep[e.g.,][]{xu2015}, the more readily available approaches suitable for most time series feature data can, broadly speaking, be categorized as based on (1) centroids, (2) distributions, (3) density, (4) hierarchies, or (5) a combination thereof \citep[see \tblref{tab:clusterApproaches} for an overview; also see][for a broader review]{jain1999}.

There is, unfortunately, no one-size-fits-all solution to clustering and users will usually have to make an informed decision based on the structure of their data as well as an appropriate weighing of accuracy and efficiency. We provide a short intuitive explanation for common approaches, together with some of their characteristics and example algorithms in \tblref{tab:clusterApproaches}. For our own illustration, we have chosen that centroid-based k-means clustering. While k-means comes at the expense of high accuracy with more complex cluster shapes, we specifically chose k-means because it is an extremely efficient method that works well with large participant- and feature numbers without making too many restrictive assumptions about the shape of the data \citep{jain2010}. K-means is also well-established within the research community and has been readily implemented in many statistical software packages \citep{hand2005}. Additionally, many of the feature selection methods have specifically been designed for the well-established k-means algorithm \citep[e.g.,][]{boutsidis2010}. As such, the k-means offers a good starting point for many psychological researchers and the method should be generalizable across a relatively wide variety of projects.

%Each of these approaches can be a valuable clustering approach for time series feature data and users will usually have to make an informed decision based on the structure of their data as well as an appropriate weighing of accuracy and efficiency. As an example, under ideal conditions, most approaches are likely to provide a similar cluster solution \citep[e.g., for well-separated groups with little noise and few outliers; e.g., see][]{peng2022}. However, when the shapes of clusters, for example, become more complex in a multi-dimensional space, density-based or hierarchy-based approaches that allow for more bottom-up clustering are likely to be more accurate \citep[e.g.,][]{langfelder2008}. Yet, with higher numbers of participants and features, both density- and hierarchy-based approaches may perform less well and the more efficient centroid-based methods might be more effective \citep[e.g.,][]{jain2010}. Similarly, several of the proposed time series features are statics that are generated from Gaussian distributions and a distribution-based algorithm might be ideally suited to separate such distributions \citep[e.g.,][]{corduas2008}. Yet, in cases where misspecifications are costly, a method with fewer assumptions might be more advisable \citep[e.g.,][]{ankerst1999}.
